Do moral communities have a spatial dimension? A spatial exploratory analysis of places of worship and violent crime in the city of Recife, Brazil

### 2020-06-11 Moral-Communities-and-Crime/Moral_Communities_and_Crime.tex
v2 Finished revisions, sent back to Edivaldo and Tatiane for comments.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Since places of worship are associated with normative codes of
  behavior, can they by their presence contribute to create moral
  communities? The mechanisms for this are explained by Groff (Groff
  2015) in terms of ``Eyes on the Street'' (see Jacobs 1961, 1968;
  Appleyard 1980) and Human Territorial Functioning (Taylor 1988, 1998).
\item
  If so, what is the distance decay of said moral community effect? Is
  it uniform over a certain distance or does it vary?
\end{enumerate}

The case study is the city of Recife in the state of Pernambuco, in
Brazil's Northeast. Recife is a large and important metropolitan area in
a historically poor region, and afflicted by high levels of violent
crime, having the dubious honor of being one of five state capitals with
the highest rates of homicide in the period under study (Menezes et al.
2013). The empirical strategy is to use spatial analysis to explore the
potential geographical relationships between violent criminal events, on
the one hand, and places of worship and a selection of commercial
establishments that serve as controls, on the other. Dissagregated data
allows us to analyze the relevant point patterns at a very high level of
resolution.

After this introductory section, the rest of this paper is structured as
follows: Section \ref{background} reviews some relevant theoretical
perspectives on religion and crime; then Section \ref{methods} presents
the empirical strategy used in the work. Section \ref{case} presents the
context as well as the data used in the study; Section \ref{results}
presents the analysis, whereas the results are discussed in Section
\ref{discussion}. Finally, some concluding remarks are offered in
Section \ref{conclusions}.

@@

The hypothesis of moral communities has over the years been used to
examine a variety of outcomes of interest. Recent examples include
Stroope and Baker's (2018) exploration of religiosity and self-rated
health; Davignon and Thomson (2015) with their research on institutional
context and the religiosity of students; and the study of religion as a
source of trust of Traunmuller (2011). This is in addition, of course,
to numerous studies on criminal behavior such as Eitle (2011), an author
who explored the deterrence power of religion on gambling; Lee and
Bartkowski's (2004) investigation of juvenile homicide in rural areas;
and the research of Regnerus (2003) on adolescent delinquency. The
hypothesis of moral communities as a form of informal social control, on
the other hand, has been less studied from a geographical perspective,
and it is only recently that has attracted the attention of researchers.
Groff (2015), for instance, discusses informal social control as a
phenomenon that can plausibly operate at different geographical scales,
from the level of the home and family, through the street block and
neighborhood, possibly to other scales such as the county. In this way,
Nie and Yang (2019) remark on a recent paper on the lack of research
conducted to study how the religious context of a geographical area
(e.g., a county) may influence (youth smoking) behavior (p.~2). This
point is echoed by Warner and Konkel (2019), who moreover recommend the
use of even smaller units of analysis (e.g., Census Block Groups), in an
effort to reduce aggregation bias (see Hipp 2007), but more importantly
because the processes defined by social disorganization
theory\footnote{The social disorganization theory (Shaw e McKay, 1942)  presents the social structure as an indispensable factor for the elaboration of theoretical and methodological models related to crime in urban neighborhoods. According to Villareal e Silva [-@Villarreal2006social], “neighborhoods characterized by dense social networks will experience greater trust among residents and cooperation in the enforcement of social norms against crime. Disadvantaged neighborhoods are thought to have fewer social ties among residents and lower rates of organizational participation”. The argument is that community characteristics such as poverty, ethnic heterogeneity and residential mobility disrupt the social organization of urban neighborhoods in such a way as to reduce their capacity to exercise social control [@Rodrigues2012medo]. Thus, aspects such as residential instability or low socioeconomic status would contribute to the weakening of the community organization, increasing crime rates. This theory highlights the importance of the geography of crime and the characteristics of disorder in its analysis.}
are thought to occur at relatively small geographical scales.

The preceding review, as well as other research (2019), makes clear the
need for research at various geographical scales. From a theoretical
perspective, the social mechanisms that underpin the hypothesis of moral
communities and informal social control can happen at various
geographical scales, some of which previous research has already
addressed: the work of Nie and Yang (2019), for one, highlights the role
of processes detectable at a relatively high level of aggregation,
whereas Warner and Konkel (2019), looking specifically at the effect of
places of worship, bring the study closer to the level of the
neighborhood. In what follows, we propose to adopt a much more
disaggregated approach by considering the intensity of criminal events
with respect to places of worship. This requires micro-level data both
of crimes and places of worship, and the use of appropriate spatial
analytical tools for point patterns, as discussed next.

@@

Brazil is a country with a Christian predominance. In the first national
census in which religion was considered, in 1872, practically the entire
population (99.72\%) identified themselves as Catholic. In the 1940
census, when the religion question was researched again, Catholicism was
still the religion with which the vast majority of the population
identified: 95.2\% of Brazilians declared themselves as Catholics that
year.

The dominance of Catholicism in the religious life of Brazil is
unsurprising given the colonial past of the country. However, the
religious makeup of the population started was not completely
homogeneous. Among other offerings in the market place of spiritual
ideas, Spiritism stands out. The Spiritist Doctrine was founded in
nineteenth-century France by Allan Kardec as a way of understanding the
world and its relations with the ``beyond'': a curious mix of
pseudo-scientific language, philosophy, and religion (Arribas 2011a).
Spiritism arrived in Brazil in the middle of the previous century and
since then it slowly spread in large and medium-sized cities. Today,
Spiritism is the third largest religious segment in Brazil, after
Catholicism and Evangelical Christianism, and shares with these
religions the Christian principle of charity through pious assistance.

Another signal event in the developmento of religion in Brazil is the
arrival of Pentecostal Protestant churches, such as the Christian
Congregation of Brazil in 1910 and the Assembly of God in 1911.
Initially, these religious denominations did not make important inroads
in Brazilian society (as evidenced by the 1940 census). However, in the
context of rapid industrialization and major urbanization, they
contributed to the change in the religious landscape along with other
religions (Souza 2019): according to the 2010 census the configuration
of the religious scene in Brazil had changed considerably from the
previous century: 64.6\% of the population identified as Catholic and
22.2\% as Evangelical Christians. The acceleration of this demographic
change was directly related to the inroads made by Evangelical churches:
The Assembly of God, for instance, grew considerably to total more than
12.3 million adherents. This church is represented extensively in many
regions of the country, following a system of loosely allied ministries.
With a charismatic profile and the lack of a centralized authority,
growth of this church has occurred mainly through subdivision into
numerous small and autonomous churches. These churches have in common
the Prosperity Gospel and an emphasis on economic entrepreneurship
(e.g., Mora 2008). They also do not shy away from partisan political
action.

In general, there have been two major changes in Brazil regarding
religious issues: the slow decline of Catholicism and the rapid growth
of Evangelical Christianism. These changes are evident in the daily life
of the average Brazilian, either through the radio or through TV
programs, which give more and more space to religious doctrines. At the
same time, there has been an increase in cases of religious intolerance
in Brazil, generating violence acts against temples, objects of worship,
and people.

@@

where \(\lambda(s)\) is the intensity of the process at point \(s\),
given by the expected number of events in a small area \(\text{d}s\)
around \(s\), as the area becomes arbitrarily small. If the process is
homogeneous (i.e., the intensity is a constant over space), an apt
estimator of the intensity is the global intensity, simply the number of
events divided by the area of the region under analysis: \[
\hat{\lambda} = \frac{n}{|\mathcal{A}|}
\]

@@

@@

The analysis is implemented using the \texttt{R} statistical computing
language, and documents to replicate the analysis, in addition to all
data necessary, can be obtained from this anonymous shared folder:

\url{https://drive.google.com/open?id=1tuJM4Mhi0Ftq3ZEwjv6RP9vO3veGjIR_}

@@

As seen in the table, Recife is a city with large socio-economic and
demographic disparities; for example, the \emph{setor} with the highest
median income has a median income that is 6,517\% higher than the median
income in the \emph{setor} with the lowest median income. Whereas there
are \emph{setores} where the proportion of population who are unemployed
is zero, there are \emph{setores} where almost 60\% of the population
are unemployed. Likewise, there are \emph{setores} where almost 70\% of
the population lives in poverty. In addition to these economic
indicators, two variables are used to represent family disruption, the
proportion of families whose head is a single mother, and the proportion
of families whose head is a \emph{young} single mother, that is, a woman
between the ages of 15 and 25. As can be seen, there are \emph{setores}
with approximately 70\% of households led by single mothers, and of
these, over 21\% are led by younger women, indicating a high degree of
family disruption.

For the analysis, we use Principal Component Analysis, a data reduction
technique, to obtain an indicator of Socio-Economic Deprivation,
essentially the first principal component in the output. The loadings of
this factor (Table \ref{tab:sed-descriptive-statistics}) indicate that
high Socio-Economic Deprivation is a combination of (in terms of
importance): high levels of poverty, high levels of unemployment, and
low median income, followed by high levels of family disruption, in
particular proportion of families led by young single mothers. This
factor accounts for almost 54\% of the variance..

@@

\caption{\label{tab:table-sed-descriptive-statistics}\label{tab:sed-descriptive-statistics}Descriptive statistics of some key socio-economic and demographic variables in Recife}
\centering
\begin{tabular}[t]{lccccccc}
\toprule
Variable & Min & 2nd Quartile & Median & Mean & 3rd Quartile & Max & \makecell[l]{PC Factor 1\\ Loadings}\\
\midrule
Population Density & 0.00 & 10564.16 & 16333.62 & 19230.23 & 25106.07 & 181642.55 & 0.03\\
Median Income & 122.05 & 337.40 & 581.67 & 1031.09 & 1283.79 & 7954.19 & -0.42\\
Proportion Unemployed & 0.00 & 0.05 & 0.10 & 0.10 & 0.14 & 0.59 & 0.56\\
Proportion Poverty & 0.00 & 0.05 & 0.11 & 0.12 & 0.17 & 0.63 & 0.58\\
Proportion Single Mother & 0.00 & 0.33 & 0.37 & 0.37 & 0.41 & 0.70 & 0.17\\
\addlinespace
Proportion Young Single Mother & 0.00 & 0.01 & 0.02 & 0.02 & 0.02 & 0.21 & 0.38\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{The variable for young single mothers is for women aged 15-25}\\
\multicolumn{8}{l}{The first principal component accounts for 53.91\% of the variance}\\
\end{tabular}
\end{table}

@@

Once the indicator of Socio-Economic Deprivation has been obtained, as
outlined above, it can be used for conditional quadrat analysis. As
previously described, the quadrats are irregular areas that are defined
based on a covariate, in this case the quintiles of the Socio-Economic
Deprivation indicator. In essence, this entails counting the number of
events at each level of the covariate, and then calculating an estimator
of the intensity. The global intensity of each type of event (i.e.,
places of worship and businesses), as well as quadrant counts and
intensity at each level of SED can be consulted in Table
\ref{tab:quadrat-count-results}.

@@

The objective of this research was to investigate the covariation of
incidence of crime to proximity of places of worship. The hypothesis of
moral communities (closely related to social disorganization theory)
posits that a negative correlation between the presence of churches and
crime. Recent papers have argued that it is important to study this
phenomenon from a geographical perspective, paying attention to the
level of aggregation. Accordingly, our research for the city of Recife
in Brazil makes the following contributions to the literature:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  By using disaggregated data for Lethal and Intentional Violent Crime
  and various types of places of worship, we were able to analyze the
  intensity of crime with respect to proximity to churches as a point
  pattern.
\item
  Following the suggestion of Warner and Konkel (2019), we also
  disaggregate our data by type of place of worship, and consider the
  three following denominations: Catholic, Evangelical, and Spiritist.
\item
  Use of relative distribution functions is effectively a multi-scale
  technique, since the intensity of crime can be estimated at different
  distances from various types of places of worship.
\item
  We also used a set of putatively morally neutral commercial
  establishments to serve as controls in our analysis.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Catholic places of worship do not seem to geographically project more
  or less of a moral community effect than, say, ice cream shops,
  pharmacies, or supermarkets: the intensity of crime as a function of
  distance to Catholic churches is not significantly different that what
  we observe when we consider distance to any of these types of
  commercial establishments, at any distance.
\item
  The intensity of crime with respect to distance to Evangelical
  churches is more complex than previously known: whereas the intensity
  of crime is higher at relatively short distance from Evangelical
  places of worship, at longer distances it tends to decline even below
  the global intensity of crime and the background level of crime
  according to Socio-Economic Deprivation.
\item
  Spiritist churches are associated with lower intensity of crime at
  very short distance; however, at longer distances(approximately
  between 200 m and 500 m), the intensity of crime is in fact higher
  than the intensity of crime at comparable distances from Catholic and
  Evangelical places of worship, or any of the three kinds of commercial
  establishments.
\end{enumerate}

These results are interesting of and in themselves as they help to
clarify the potential of various types of moral communities to act as
restraints on crime or, on the contrary, as criminogenic factors. They
also suggest several avenues for further research.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The analysis presented in this paper is to a large extent exploratory.
  For example, to introduce a baseline function in the analysis of
  relative distribution, we used a data reduction technique to obtain an
  indicator of Socio-Economic Deprivation. This indicator was useful to
  determine a general spatial pattern of crime conditional on this
  variable, and to account for a baseline (or background) intensity of
  crime. However, it would be interesting to investigate the effect of
  the individual socio-economic and demographic variables on the
  intensity of crime, as opposed to the aggregated effect of
  Socio-Economic Deprivation. This implies the use of multivariate
  analytical techniques.
\item
  It would be interesting to explore the effect of aggregation on the
  results. On the one hand, this would be informative with respect to
  the well-known issues with aggregation bias in spatial analysis
  (related to the Modifiable Areal Unit Problem in geography; see
  Fotheringham and Wong 1991; Openshaw and Taylor 1979; and Tagashira
  and Okabe 2002). On the other hand, as the results of the research
  presented here, it is possible that analysis at different levels of
  aggregation may tend to capture different spatial and social processes
  of interest.
\item
  And finally, in relation to the latter point, more in-depth research,
  perhaps observational, ethnographic, or participatory studies, could
  help to develop a more refined understanding of the kinds of processes
  that operate at different scales and how they can influence moral
  behavior and crime.
\end{enumerate}

@@

@@

@@

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Since places of worship are associated with normative codes of
  behavior, can they by their presence contribute to create moral
  communities? The mechanisms for this are explained by Groff (Groff
  2015) in terms of ``Eyes on the Street'' (see Jacobs 1961, 1968;
  Appleyard 1980) and Human Territorial Functioning (Taylor 1988, 1998).
\item
  If so, what is the distance decay of said moral community effect? Is
  it uniform over a certain distance, such that it could be well
  approximated by areas, or does it vary?
\end{enumerate}

The case study is the city of Recife in the state of Pernambuco, in
Brazil's Northeast. Recife is a large and important metropolitan area in
a historically poor region, and afflicted by high levels of violent
crime, having the dubious honor of being one of five state capitals with
the highest rates of homicide in the period under study (Menezes et al.
2013). The empirical strategy is to use spatial analysis to explore the
potential geographical relationships between violent criminal events, on
the one hand, and places of worship and a selection of commercial
establishments that serve as controls, on the other.

Note that this is a reproducible research document. The analysis is
implemented using the \texttt{R} statistical computing language, and
documents to replicate the analysis, in addition to all data necessary,
are available from a
repository\footnote{\url{https://drive.google.com/open?id=1tuJM4Mhi0Ftq3ZEwjv6RP9vO3veGjIR_}}.

@@

The hypothesis of moral communities has over the years been used to
examine a variety of outcomes of interest. Recent examples include
Stroope and Baker's (2018) exploration of religiosity and self-rated
health; Davignon and Thomson (2015) with their research on institutional
context and the religiosity of students; and the study of religion as a
source of trust of Traunmuller (2011). This is in addition, of course,
to numerous studies on criminal behavior such as Eitle (2011), an author
who explored the deterrence power of religion on gambling; Lee and
Bartkowski's (2004) investigation of juvenile homicide in rural areas;
and the research of Regnerus (2003) on adolescent delinquency. The
hypothesis of moral communities as a form of informal social control, on
the other hand, has been less studied from a geographical perspective,
and it is only recently that has attracted the attention of researchers.
Groff (2015), for instance, discusses informal social control as a
phenomenon that can plausibly operate at different geographical scales,
from the level of the home and family, through the street block and
neighborhood, possibly to other scales such as the county. In this way,
Nie and Yang (2019) remark on a recent paper on the lack of research
conducted to study how the religious context of a geographical area
(e.g., a county) may influence (youth smoking) behavior (p.~2). This
point is echoed by Warner and Konkel (2019), who moreover recommend the
use of smaller units of analysis (e.g., Census Block Groups), in an
effort to reduce aggregation bias (see Hipp 2007), but more importantly
because the processes defined by social disorganization
theory\footnote{The social disorganization theory (Shaw e McKay, 1942)  presents the social structure as an indispensable factor for the elaboration of theoretical and methodological models related to crime in urban neighborhoods. According to Villareal e Silva [-@Villarreal2006social], “neighborhoods characterized by dense social networks will experience greater trust among residents and cooperation in the enforcement of social norms against crime. Disadvantaged neighborhoods are thought to have fewer social ties among residents and lower rates of organizational participation”. The argument is that community characteristics such as poverty, ethnic heterogeneity and residential mobility disrupt the social organization of urban neighborhoods in such a way as to reduce their capacity to exercise social control [@Rodrigues2012medo]. Thus, aspects such as residential instability or low socioeconomic status would contribute to the weakening of the community organization, increasing crime rates. This theory highlights the importance of the geography of crime and the characteristics of disorder in its analysis.}
are thought to occur at relatively small geographical scales.

Groff (Groff 2015), in particular, explains the small-scale mechanisms
that could be in operation in terms of ``Eyes on the Street'' and Human
Territorial Functioning. ``Eyes on the Street'' (see Jacobs 1961, 1968;
Appleyard 1980) refers to involvment in small civic acts by users of
public spaces, which signal social organization to an observer.
Community engagement, moreover, is engendered at small scales, as shown
by authors that include Appleyard (1980) and more recently Grannis
(2009). Human Territorial Functioning (HTF; Taylor 1988, 1998) also
provides a valuable perspective on social informal control This concept
posits that territorial functioning is a product of the shared
perceptions of individuals in regards to places, and how these
perceptions are shaped by the physical, social, and cultural
environments. For example, Paez (2013) discusses how perceptions of
safety are the outcome of a socially interactive process, and shows how
this process leads to spatial patterning at very small-scales. The
preceding review, as well as other research (2019), makes clear the need
for research at various geographical scales. From a theoretical
perspective, the social mechanisms that underpin the hypothesis of moral
communities and informal social control can happen at various
geographical scales, some of which previous research has already
addressed: the work of Nie and Yang (2019), for one, highlights the role
of processes detectable at a relatively high level of aggregation,
whereas Warner and Konkel (2019), looking specifically at the effect of
places of worship, bring the study closer to the level of the
neighborhood.

@@

Brazil is a predominantly Christian country. The first national census
that collected information on religion was in 1872, a year when
practically the entire population (99.72\%) identified themselves as
Catholic. In the 1940 census, when the religion question was asked
again, Catholicism was still the dominant religion with 95.2\% of
Brazilians declaring themselves as Catholics. The dominance of
Catholicism in the religious life of Brazil is unsurprising given the
country's past as a former colony of a Catholic power (Portugal).
However, the religious makeup of the population started to become less
homogeneous in the second half of the twentieth century. Among other
offerings in the market place of spiritual ideas, Spiritism stands out.
The Spiritist Doctrine was founded in nineteenth-century France by Allan
Kardec as a way of understanding the world and its relations with the
``beyond'', in a curious mix of pseudo-scientific language, philosophy,
and religion (Arribas 2011a). Spiritism arrived in Brazil in the middle
of the nineteenth century, and since then it slowly spread in large and
medium-sized cities. Today, Spiritism is the third largest religious
segment in Brazil, after Catholicism and Evangelical Christianism, and
shares with these religions the Christian principle of charity through
pious assistance.

Another signal event in the developmento of religion in Brazil was the
arrival of Pentecostal Protestant churches, such as the Christian
Congregation of Brazil in 1910 and the Assembly of God in 1911.
Initially, these religious denominations did not make important inroads
in Brazilian society (as evidenced by the 1940 census). However, in the
context of rapid industrialization and major urbanization, they
contributed to the change in the religious landscape along with other
religions (Souza 2019): according to the 2010 census the configuration
of the religious scene in Brazil had changed considerably from the
previous century: 64.6\% of the population identified as Catholic and
22.2\% as Evangelical Christians. The acceleration of this demographic
change was directly related to successful proselitism by Evangelical
churches: The Assembly of God, for instance, is a church that grew
considerably to total more than 12.3 million adherents. This church is
represented extensively in many regions of the country, following a
system of loosely allied ministries. With a charismatic profile and the
lack of a centralized authority, growth of this church has occurred
mainly through subdivision into numerous small and autonomous churches.
These churches have in common the Prosperity Gospel and an emphasis on
economic entrepreneurship (e.g., Mora 2008). They also do not shy away
from partisan political action.

In general, there have been two major changes in Brazil regarding
religious issues: the slow decline of Catholicism and the rapid growth
of Evangelical Christianism. These changes are evident in the daily life
of the average Brazilian, either through the radio or through TV
programs, which give more and more space to religious doctrines. At the
same time, there has been an increase in cases of religious intolerance
in Brazil, generating violence acts against temples, objects of worship,
and people (e.g., Souza 2019; Fonseca 2018).

In what follows, we propose to adopt a disaggregated approach by
considering the intensity of criminal events with respect to places of
worship. This requires micro-level data both of crimes and places of
worship, and the use of appropriate spatial analytical tools for point
patterns, as discussed next.

@@

\noindent where \(\lambda(s)\) is the intensity of the process at point
\(s\), given by the expected number of events in a small area
\(\text{d}s\) around \(s\), as the area becomes arbitrarily small. If
the process is homogeneous (i.e., the intensity is a constant over
space), an apt estimator of the intensity is the global intensity,
simply the number of events divided by the area of the region under
analysis: \[
\hat{\lambda} = \frac{n}{|\mathcal{A}|}
\]

@@

\hypertarget{interpretation}{%
\subsection{Interpretation}\label{interpretation}}

Both quadrats and the relative distribution function are estimators of
intensity. In the case of quadrats, the estimator is the number of
events in the quadrat divided by area of the quadrat (conditional
quadrat counts use irregular quadrats that are defined based on some
control variable, in our case SED). In the case of the relative
distribution, the output is an estimate of intensity conditioned on
distance from a signal event, e.g., a place of worship. The relative
distribution can be adjusted by means of a baseline function \(B(s)\),
which is also an estimate of intensity but conditioned on some control
covariate (in our cases SED). Therefore, the interpretation of the
relative distribution with a baseline function is as a ratio of the
intensity with respect to distance, to the baseline intensity according
to the control covariate. If the ratio, which is dimensionless, is 1,
then intensity as a function of distance is equal to the baseline
intensity.

@@

@@

@@

\caption{\label{tab:table-sed-descriptive-statistics}\label{tab:sed-descriptive-statistics}Descriptive statistics of some key socio-economic and demographic variables in Recife}
\centering
\begin{tabular}[t]{lccccccc}
\toprule
Variable & Min & 2nd Quartile & Median & Mean & 3rd Quartile & Max & \makecell[l]{PC Factor 1\\ Loadings}\\
\midrule
Median Income & 122.05 & 337.40 & 581.67 & 1031.09 & 1283.79 & 181642.5 & -0.42\\
Proportion Unemployed & 0.00 & 0.05 & 0.10 & 0.10 & 0.14 & 181642.5 & 0.56\\
Proportion Poverty & 0.00 & 0.05 & 0.11 & 0.12 & 0.17 & 181642.5 & 0.58\\
Proportion Single Mother & 0.00 & 0.33 & 0.37 & 0.37 & 0.41 & 181642.5 & 0.17\\
Proportion Young Single Mother & 0.00 & 0.01 & 0.02 & 0.02 & 0.02 & 181642.5 & 0.38\\
\addlinespace
Population Density & 0.00 & 10564.16 & 16333.62 & 19230.23 & 25106.07 & 181642.5 & 0.03\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{The variable for young single mothers is for women aged 15-25}\\
\multicolumn{8}{l}{The first principal component accounts for 44.9\% of the variance}\\
\end{tabular}
\end{table}

As seen in the table, Recife is a city with large socio-economic and
demographic disparities; for example, the \emph{setor} with the highest
median income has a median income that is 148,826\% higher than the
median income in the \emph{setor} with the lowest median income. Whereas
there are \emph{setores} where the proportion of population who are
unemployed is zero, there are \emph{setores} where almost 60\% of the
population are unemployed. Likewise, there are \emph{setores} where
almost 70\% of the population lives in poverty. In addition to these
economic indicators, two variables are used to represent family
disruption, the proportion of families whose head is a single mother,
and the proportion of families whose head is a \emph{young} single
mother, that is, a woman between the ages of 15 and 25. As can be seen,
there are \emph{setores} with approximately 70\% of households led by
single mothers, and of these, over 21\% are led by younger women,
indicating a high degree of family disruption. Unlike other places,
where higher population density is associated with more poverty, the
distribution of this variable in Recife is more mixed: middle income and
higher income households favor high density development in many parts of
the city. The spatial distribution of these variables can be seen in
Figure \ref{fig:plot-sed-variables}.

\includegraphics{Moral_Communities_and_Crime_files/figure-latex/fig-plot-sed-variables-1} \hfill{}

\caption{\label{fig:plot-sed-variables}Spatial distribution of socio-economic and demographic variables in Recife}\label{fig:fig-plot-sed-variables}
\end{figure}

For the analysis, we use Principal Component Analysis, a data reduction
technique, to obtain an indicator of Socio-Economic Deprivation,
essentially the first principal component in the output. The loadings of
this factor (Table \ref{tab:sed-descriptive-statistics}) indicate that
high Socio-Economic Deprivation is a combination of (in terms of
importance): high levels of poverty, high levels of unemployment, and
low median income, followed by high levels of family disruption, in
particular proportion of families led by young single mothers. This
factor accounts for almost 45\% of the variance.

@@

Once the indicator of Socio-Economic Deprivation has been obtained, as
outlined above, it can be used for conditional quadrat analysis. As
previously described, the quadrats are irregular areas that are defined
based on a covariate, in this case the quintiles of the Socio-Economic
Deprivation indicator. Each color in Figure
\ref{fig:plot-sed-as-quintiles} is a ``container'' for counting events.
In essence, this entails counting the number of events at each level of
the covariate, and then calculating an estimator of the intensity. The
global intensity of each type of event (i.e., places of worship and
businesses), as well as quadrant counts and intensity at each level of
SED can be consulted in Table \ref{tab:quadrat-count-results}.

@@

The objective of this research was to investigate the covariation of
incidence of crime to proximity of places of worship. The hypothesis of
moral communities (closely related to social informal controls) posits a
negative correlation between the presence of churches and crime. Recent
papers have argued that it is important to study this phenomenon from a
geographical perspective, paying attention to the level of aggregation.
Accordingly, our research for the city of Recife in Brazil makes the
following contributions to the literature:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  By using disaggregated data for Lethal and Intentional Violent Crime
  and various types of places of worship, we were able to analyze the
  intensity of crime with respect to proximity to churches as a point
  pattern.
\item
  Following the suggestion of Warner and Konkel (2019), we also
  disaggregated our data by type of place of worship, and considered the
  three following denominations: Catholic, Evangelical, and Spiritist.
\item
  We used relative distribution functions, a multi-scale technique, that
  allowed us to estimate the intensity of crime at different distances
  from various types of places of worship.
\item
  We also used a set of putatively morally neutral commercial
  establishments to serve as controls in our analysis.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Catholic places of worship do not seem to geographically project more
  or less of a moral community effect than, say, ice cream shops,
  pharmacies, or supermarkets: the intensity of crime as a function of
  distance to Catholic churches is not significantly different that what
  we observe when we consider distance to any of these types of
  commercial establishments, at any distance.
\item
  The intensity of crime with respect to distance to Evangelical
  churches is more complex than previously known: whereas the intensity
  of crime is higher at relatively short distance from Evangelical
  places of worship, at longer distances it tends to decline even below
  the global intensity of crime and the background level of crime
  according to Socio-Economic Deprivation.
\item
  Spiritist churches are associated with lower intensity of crime at
  very short distance; however, at longer distances (approximately
  between 200 m and 500 m), the intensity of crime is in fact higher
  than the intensity of crime at comparable distances from Catholic and
  Evangelical places of worship, or any of the three kinds of commercial
  establishments.
\end{enumerate}

These results are interesting in and of themselves as they help to
clarify the potential of various types of moral communities to act as
restraints on crime or, on the contrary, as criminogenic factors. They
also suggest several recommendations.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The analysis presented in this paper is to a large extent exploratory.
  For example, to introduce a baseline function in the analysis of
  relative distribution, we used a data reduction technique to obtain an
  indicator of Socio-Economic Deprivation. This indicator was useful to
  determine a general spatial pattern of crime conditional on this
  variable, and to account for a baseline (or background) intensity of
  crime. However, it would be interesting to investigate the effect of
  the individual socio-economic and demographic variables on the
  intensity of crime, as opposed to the aggregated effect of
  Socio-Economic Deprivation. This implies the use of multivariate
  analytical techniques.
\item
  It would be interesting to explore the effect of aggregation on the
  results. On the one hand, this would be informative with respect to
  the well-known issues with aggregation bias in spatial analysis
  (related to the Modifiable Areal Unit Problem in geography; see
  Fotheringham and Wong 1991; Openshaw and Taylor 1979; and Tagashira
  and Okabe 2002). On the other hand, as the results of the research
  presented here, it is possible that analysis at different levels of
  aggregation may tend to capture different spatial and social processes
  of interest.
\item
  And finally, in relation to the latter point, more in-depth research,
  perhaps observational, ethnographic, or participatory studies, could
  help to develop a more refined understanding of the kinds of processes
  that operate at different scales and how they can influence moral
  behavior and crime. For example, an important question is whether the
  higher intensity of crime in the neighborhood of Evangelical churches
  is related to ``The Violence of the Just'' (see Birman and Machado
  2012; Birman 2019), or contrariwise, crime committed \emph{against}
  Evangelical Christians as a result of religious intolerance (see Souza
  2019; Fonseca 2018).
\end{enumerate}

@@

\leavevmode\hypertarget{ref-Fonseca2018primeiras}{}%
Fonseca, Alexandre Brasil. 2018. ``Primeiras Análises Dos Dados Do
Relatório Sobre Intolerância E Violência Religiosa No Brasil
(2011-2015).'' Journal Article. \emph{Estado Laico, Intolerância E
Diversidade Religiosa No Brasil. Brasília, Ministério Dos Direitos
Humanos}, 22--47.

@@

\leavevmode\hypertarget{ref-Grannis2009from}{}%
Grannis, R. 2009. \emph{From the Ground up: Translating Geography into
Community Through Neighbor Networks}. Book. Princeton: Princeton
University Press.

@@

\leavevmode\hypertarget{ref-Paez2013mapping}{}%
Paez, A. 2013. ``Mapping Travelers' Attitudes: Does Space Matter?''
Journal Article. \emph{Journal of Transport Geography} 26: 117--25.
\url{https://doi.org/10.1016/j.jtrangeo.2012.09.002}.
